### 2009-12-02 docs/marcello-dissertation/thesis/chapter1-introduction.tex
Revision after kleber's feedback on first and part of the second chapters.
git-svn-id: svn://localhost/netbeams/branches/marcello/persistence@666 559edde2-a4d4-44e0-96fc-4035e40d7396

Sensor networks are commonly used in different areas for different purposes
such as accurate measurements and the general public use \cite{sn-intro01}
\cite{sn-intro02}. In the scientific community, they serve as tools to watch
the state of the environment by storing samples of data at particular periods
of time. For example, the NASA Jet Propulsion Laboratory uses the Volcano
SensorWeb \cite{sn-ex02}, a sensor network which aims at collecting data from
individual volcanoes in Alaska and South Pacific to be used with heuristics
regarding hazardous activities. Similarly, the National Data Buoy Center, a
division of the National Oceanic and Atmospheric Administration (NOAA)
\cite{sn-ex03}, provides online information from different buoys in different
shores around the world regarding water quality, current information, etc,
what characterizes sensor network for environmental monitoring \cite{sn-ex01}.
Therefore, sensor networks can be seen as an extremely valuable scientific
research and commercial sector with direct application and use of its produced
data.

In order to provide access to the collected sensor data, different approaches
can be taken to interrogate sensor devices indirectly by accessing its data in
places called network sink. Similarly, the data model and database system
location may also influence in the process of reusing the collected data. In
this way, the lack of a persistence layer for existing sensor networks requires
a deep understanding about its infrastructure, the nature and location of the
collected data, the chosen data model representation, among others. For
instance, how can a data persistence layer safely represent the collected
sensor devices' properties without the necessity of restructuring any existing
data model? Should the end users' skills in database system be taken into
account or the use of programming languages to access the data
\cite{sn-programming-language}? This problem is currently faced by the
department of Biology at San Francisco State University with a marine sensor
network managed by researchers without Computer Science skills. While the data
access may enforce the difficulties, technical issues may also increase while
the implementation of a sensor network such as growth in terms of the number of
sensor devices and, consequently, the number of data produced. Last, but not
least, the way to read the data for reuse is the most striking features of the
sensor network, taking into account the different data formats used by the
primary users of the system. How can individual collected sensor data be
exported to other formats such as spreadsheets or archived files meant to be
shared among research peers? For these reasons, the inception of a persistence
storage can be an exceptionally challenging problem to solve.

>>>>

Sensor networks are commonly used in different areas for different purposes
such as accurate measurements during scientific research \cite{sn-intro01} and
the general public access such as weather forecast \cite{sn-intro02}. In the
scientific community, they serve as tools to watch the state of the environment
by storing samples of data at particular periods of time. For example, the
NASA Jet Propulsion Laboratory uses the Volcano SensorWeb \cite{sn-ex02}, a
sensor network which aims at collecting data from individual volcanoes in
Alaska and South Pacific to be used with heuristics regarding hazardous
activities. Similarly, the National Data Buoy Center, a division of the
National Oceanic and Atmospheric Administration (NOAA) \cite{sn-ex03},
provides online information from different buoys in different shores around
the world regarding water quality, current information, etc, what
characterizes sensor network for environmental monitoring \cite{sn-ex01}.
Therefore, sensor networks can be seen as an extremely valuable tool
for the scientific and commercial sectors, makeing direct use of the sensor
networks' produced data.

In order to provide access to the collected sensor data, different approaches
can be taken: interrogate sensor devices indirectly by accessing its data in
places called network data sink or simply connect directly to the device. The 
While the latter approach the data is only available in memory as a data
stream, the former approach contains any sort of secondary memory such as a
hard-disk or a flash memory, which are used to archive the collected data.
Along with the location of the collected data from sensors, the the data model
and may also influence in the process of reusing the collected data. In this
way, the lack of a persistence layer for existing sensor networks requires a
deep understanding about its infrastructure, the nature and location of the
collected data, the chosen data model representation, among others. For
instance, how can a data persistence layer safely represent the collected
sensor devices' properties without the necessity of restructuring any existing
data model? Should the end users' skills in database system be taken into
account or the use of programming languages to access the data
\cite{sn-programming-language}? This problem is currently faced by the
department of Biology at San Francisco State University with a marine sensor
network managed by researchers without Computer Science skills. While the data
access may enforce the difficulties, technical issues may also increase while
the implementation of a sensor network such as growth in terms of the number
of sensor devices and, consequently, the number of data produced. Last, but
not least, the way to read the data for reuse is the most striking features of
the sensor network, taking into account the different data formats used by the
primary users of the system. How can individual collected sensor data be
exported to other formats such as spreadsheets in order to be shared among
research peers? For these reasons, the inception of a persistence storage can
be an exceptionally challenging problem to solve.

### 2009-12-02 docs/marcello-dissertation/thesis/chapter10-appendix.tex
Revision after kleber's feedback on first and part of the second chapters.
git-svn-id: svn://localhost/netbeams/branches/marcello/persistence@666 559edde2-a4d4-44e0-96fc-4035e40d7396

\chapter{DSP Data Persistence: Component Implementation}
\label{sec:dsp-data-persistence-implementation}

@@

\section{DSP Platform Deployment}

@@

\begin{figure}[h]
  \centering
  \includegraphics[scale=0.3]{../diagrams/view-collections-instance-browser-futondb}
  \caption{Viewing a partial list of data using the Futon for CouchDB/MongoDB}
  \label{fig:view-collections-instance-browser-futondb}
\end{figure}

@@

\begin{figure}[h]
  \centering
  \includegraphics[scale=0.3]{../diagrams/view-collected-data-list-browser-futondb}
  \caption{Viewing a partial list of data using the Futon for CouchDB/MongoDB}
  \label{fig:view-collected-data-list-browser-futondb}
\end{figure}

@@

\begin{figure}[h]
  \centering
  \includegraphics[scale=0.3]{../diagrams/view-collected-data-instance-browser-futondb}
  \caption{Viewing an instance of collected data using the Futon for
  CouchDB/MongoDB}
  \label{fig:view-collected-data-instance-browser-futondb}
\end{figure}

@@

\lstinputlisting[language=Java,label=file:experiment-dsp-implementation,caption=Inserting
Random Data to MongoDB]{../../../../netbeams/versions/v2/apps/osgi-bundles/dsp/DSPDataPersistence/src/org/netbeams/dsp/persistence/controller/DSPMessageToMongoDBExperiment.java}

>>>>

\section{DSP Data Persistence: Component Implementation}
\label{sec:dsp-data-persistence-implementation}

@@

\section{DSP Platform Deployment Details}

@@

\begin{figure}[h]
  \centering
  \includegraphics[scale=0.5]{../diagrams/view-collections-instance-browser-futondb}
  \caption{Viewing a partial list of data using the Futon for CouchDB/MongoDB}
  \label{fig:view-collections-instance-browser-futondb}
\end{figure}

@@

\begin{figure}[h]
  \centering
  \includegraphics[scale=0.5]{../diagrams/view-collected-data-list-browser-futondb}
  \caption{Viewing a partial list of data using the Futon for CouchDB/MongoDB}
  \label{fig:view-collected-data-list-browser-futondb}
\end{figure}

@@

\begin{figure}[h]
  \centering
  \includegraphics[scale=0.5]{../diagrams/view-collected-data-instance-browser-futondb}
  \caption{Viewing an instance of collected data using the Futon for
  CouchDB/MongoDB}
  \label{fig:view-collected-data-instance-browser-futondb}
\end{figure}

@@

\lstinputlisting[language=Java,label=file:experiment-dsp-java-executor,caption=Inserting
Random Data to MongoDB]{../../../../netbeams/versions/v2/apps/osgi-bundles/dsp/DSPDataPersistence/src/org/netbeams/dsp/persistence/controller/DSPMessageToMongoDBExperiment.java}

### 2009-12-02 docs/marcello-dissertation/thesis/chapter2-state-of-the-art.tex
Revision after kleber's feedback on first and part of the second chapters.
git-svn-id: svn://localhost/netbeams/branches/marcello/persistence@666 559edde2-a4d4-44e0-96fc-4035e40d7396

Different aspects related to two fields of Computer Science have motivated to
the commencement of this work. In the prospect of Database Systems and
Distributed Systems, this work seeks a different approach for a persisting
collected data from sensor networks into a storage system that can be seen as a
cluster of distributed databases. Likewise, in the perspective of Software
Engineering, this work adds a crucial capability to an existing sensor network
framework, meeting the requirements and interfaces. In general, this chapter
gives the foundation based on the motivations for the establishment of a
different approach to persisting data using a different data model, covering
topics that will be referred throughout the following other chapters.

This chapter will first give an overview on different related problems
described in the literature that solved problems related to data modeling and
representation in sensor networks.

\section{Survey in Sensor Networks}

Sensor Networks are specialized types of network systems comprised of nodes
with specific goals to monitor and collect data. Data Persistence in Sensor
Networks, however, depends on the approaches used to store the data, since data
nodes can be located close to the devices in the network or in a centralized
node. In this way, the query process can take place on the network nodes
themselves or in the centralized data sink. The former approach is used when
the data is collected at the sensor node on its local storage
\cite{sn-storage01}\cite{sn-storage03}, strategy typically used to maximize
the sensor's resources while minimizing energy consumption
\cite{sn-storage04}. However, this strategy is used when data archival is not
the main purpose, given the limited resources on the sensor devices. For this
reason, the latter strategy has been used when the collected data is generally
archived in a centralized database system for later reuse \cite{sn-storage02}.

When it comes to data representation, the relational data model
\cite{relational-model} is one of the most used approaches to represent data in
sensor network databases in any of the query processing strategies and, in most
cases, with a modified version of the SQL language \cite{sn-db-newop}. However,
the design of the database system must be done with prior knowledge of the
sensor types when choosing the relational data model, as the database design
must be normalized \cite{db-normalization} for the entities already identified
for the system. For example, the introduction of a new sensor device might
represent a potential change on the database structure by seeking a new
normalized version of the database design. In addition to challenges of
maintaining the database schema normalized, the use of Data Provenance
approaches in sensor network data \cite{sn-provenance} usually addresses the
problem related with the origin of data and its entire history. However, once
the data model is chosen and implemented, one of the most prominent
nonfunctional requirements in sensor networks is scalability of the persistence
storage. Since an operational sensor network can generate enormous amounts of
data, it must be able to cope with the increase and data load during data
gathering, and yet provide users the same performance during data retrieval by
using different strategies such a data-centric storage \cite{sn-storage03}.

The following subsections describe classes of characteristis of these types of
sensor networks, with applications on research, industrial or militery.

The following subsections describe classes of characteristics of these types of
sensor networks, with applications on research, commercial or military.

@@

Each of the nodes can be deployed in a arbitrary method
such as being dropped from an airplane, or in a manual mode at carefully chosen
geographical points \cite{sn-intro01}. Moreover, the deployment can be done like a
one-time event or an iterative process.

However, nodes can be able to move around the deployed locality or remain fixed
in this place. In the case of the amount of the former behaviour, a sensor node
can move occasionally or continuously, being a result of an accidental
side-effect or a desired property of the device, where they are an automotive
device or attached to another moving object. For this reason, \cite{sn-intro01}
describes those types of mobility as active or passive.

@@

Such sensor nodes can be of different sizes, ranging from microscopically small
particles, small devices to bigger ones with a few feet cubic. In addition to
being limited in size, they are also limited in resources and computational
power, forcing their  conception limited to maintain its energy stored  either
on batteries or scavenged from solar cells \cite{sn-intro01}.

Since sensor devices range in size and capabilities, the cost for each of the
devices very from hundreds to thousands of dollars, depending on its
functionality, energy utilization, etc. For this reason, \cite{sn-intro01}
explains that a sensor network may potentially contain similar sensor devices,
as they can contain dissimilar ones from each other, which characterizes the
heterogeneity of a sensor network by the hardware capabilities such as GPS
enabled or having GSM network connectivity.

\subsection{Communication Modality, Network Infrastructure, Coverage and
Connectivity}
\label{sec:sn-infrastructure}

Sensor nodes can communicate in different modalities such as radio, diffuse
light, inductive or sound, which may influence on the type of protocols used to
communicate among the nodes, where they can be said to be in an organized
infrastructure or as an ad hoc way \cite{sn-intro01}.

When it comes to the sensor network different organizational infrastructure,
they can differ in types of topology: single-hop, star, networked stars, tree
and graph, where each of them may affect the characteristics of the network as
it may also influence on data collection and routing mechanisms. While in a
single-hop sensor network there are only two levels of sensor connectivity, the
tree, graph or networked stars networks uses communications among the network
hosts for data communication.

Last, but not least, sensor networks may be influenced its coverage designed in
its purpose, its size in terms of the number of nodes and its relating
lifetime. As sensors are deployed in different areas, they can be distributed
in a sparse or dense nature, yet providing essential coverage for the
environment. However, considering that sensor devices are very likely to
regular failures, projects have provided redundant nodes \cite{sn-intro01}.
This level of coverage also influences on the connectivity of the sensor
nodes, where they can be considered connected when a connection is available
at all times. On the other hand, nodes can communicate with other nodes in a
sporadic way. In the former, the sensor node is partitioned, while in the
latter the nodes are isolated most of the time and communicates at certain
moments in time.

@@

Similar to the degree of coverage of the sensor network is the number of
devices deployed in the network, where they can be in the order of dozens,
hundreds or even thousand of nodes \cite{sn-intro01}.

The application of the sensor network determines how long the sensor network
may be exist. For example, the existence of an environmental sensor networks
\cite{sn-ex01} may be not determined since its use serves in a regular basis,
whereas the lifetime of a sensor network to determine the number of people in
an event is limited to the event.

@@

\subsection{The Purpose of the Collected Data}
\label{sec:sn-data-purpose}

@@

Under these circumstances, the collected data needs first to face its journey
starting from its transmission from its producing device to the specified data
sink as specified by \cite{sn-storage01}\cite{sn-storage02}\cite{sn-storage03}.

@@

It is clear that the data flows from the sensor node to another depending on
the definition of the network. On the whole, the sensor network infrastructure
dictates where the data will flow to, as described in section
\ref{sec:sn-infrastructure}. \cite{sn-storage02} sees a wireless sensor
networks as a many-to-one data collection pattern, when its organization is
based on single-hop or star. On the other hand, when the data organization is
structured as a graph, tree, etc, the data flows from one intermediate sensor
node to another until it reaches its final storage node as described by
\cite{sn-storage01}\cite{sn-storage03}.

@@

In order to mitigate the problems related flooded nodes in in-network query
processing, the use of a Data-Centric storage strategy has been shown to be a
proven alternative for the purpose of energy consumption
\cite{sn-storage03}\cite{sn-storage01}.Its primary goal is to organize the
collected data based on a particular property, say location or event, in order
to be sent to individual storage nodes that relate to the observation. As the
main outcome of this strategy is to create network partitioning strategy, the
query processor would hit specific sensor nodes with the needed data and,
therefore, would decrease the load and maximize the capacity on the actual
sensor devices.

@@

\begin{itemize}
  \item \textbf{Tabular Data Model}: comma-separated values are immensely
  popular due to its simplicity. Although data is not index , it is an easy
  approach for data exchange among researchers \cite{sn-provenance};
  \item \textbf{Relational Data Model}: it is used in the majority of the
  projects. Most of them due to the easy access to database systems, with
  refactored versions that support the implementation of sensor networks and
  its requirements \cite{sn-db-tinydb};
  \item \textbf{Structured Data Model}: similar to the relational model,
  structured data models are becoming popular due to the Internet. The use of
  XML data \cite{xml} is not yet popular, but there are a few projects using
  it \cite{sn-xml-usage01}\cite{sn-xml-usage02}, implementing its query by
  using a middleware \cite{sn-xml-middleware}\cite{sn-xml-query-engines}.
\end{itemize}

@@

\subsection{Relational Data Model and Query}

@@

\subsection{Structured Data Model and Query}

@@

\section{Data Provenance: How to describe collected data?}
\label{sec:sn-provenance}

>>>>

In general, this chapter reviews sensor networks characteristics and
properties, necessary to the understanding of the basic properties of a sensor
network. Moreover, it reviews the characteristics of data persistence for
sensor networks, characterizing the different approaches used to store data
such as data model used and how the collected data is described.

\section{Survey in Sensor Networks and Its Data}

Sensor Networks are specialized types of network systems comprised of nodes
with specific goals such as to produce and collect data. The former is seen
as a sensor device capable of producing data to the network based on an
observation from the environment or just a raw value regarding a given
propertiy, while the latter can be seen as the network sink responsible to
receive the observed or ``sensed'' data. In this way, data persistence in sensor
networks depends on the approaches used to store the data on the data sink,
since these types of nodes can be located in the network close to the devices 
or in a centralized location. As a consequence, the query used to query the
data can take place in the network nodes themselves or in the centralized node.
The former approach is used when the data is collected at the sensor node on
its local storage \cite{sn-storage01, sn-storage03}, strategy typically
used to maximize the sensor's resources, while minimizing energy consumption
\cite{sn-storage04}.The latter strategy, however, has been used when the
collected data is archived in a centralized database system for later reuse
\cite{sn-storage02}.

When it comes to data representation, the relational data model
\cite{relational-model} is one of the most used approaches to represent data in
sensor network databases in any of the query processing strategies and, in most
cases, with a modified version of the SQL language \cite{sn-db-newop}. However,
the design of the database system must be done with prior knowledge of the
sensor types when choosing the relational data model, as the database design
must be normalized \cite{db-normalization} for the entities already identified
for the system. For example, the introduction of a new sensor device might
represent a potential change on the database structure by seeking a new
normalized version of the database design. In addition to challenges of
maintaining the database schema normalized, the use of Data Provenance 
approaches in sensor network data usually addresses the problem related with
the origin of data and its entire history \cite{sn-provenance}.

The following subsections describe the main characteristics of sensor networks
described in the literature, and then others that interest data persistence in
sensor networks.

@@

Each node of the sensor network can be deployed in an arbitrary method such as
being dropped from an airplane, or in a manually placed at chosen geographic
locations \cite{sn-intro01}. Moreover, the deployment can be done as a one-time
event or as an iterative process, depending on the needs of where the sensor
device needs to be placed.

Nodes can be able to move around the deployed locality or remain fixed at a
single place. In the case of a sensor networks comprised with different moving
sensors, a given sensor node can move occasionally or continuously, as a result
of an accidental side-effect or a desired property of the device. Usually,
these sensors are automotive or they can be attached to another moving object.
For this reason, \cite{sn-intro01} describes those types of mobility as active
or passive.

@@

Such sensor nodes can be of different sizes, ranging from microscopically small
particles, small devices to bigger ones with a few feet cubic. In addition to
being limited in size, they are also limited in resources and computational
power, as they are designed to either stored energy on batteries or
use energy scavenged from solar cells \cite{sn-intro01}.

Sensor devices range in size, capabilities and other properties, which are
associated to their cost ranging from hundreds to thousands of dollars. All in
all, the price depends on the device's functionality, energy utilization
capabilities, among others. For this reason, \cite{sn-intro01} explains that a
sensor network may potentially contain similar sensor devices, as they can
contain dissimilar ones from each other, which characterizes the heterogeneity
of a sensor network by the hardware capabilities such as GPS enabled or having
GSM network connectivity.

\subsection{Communication Mode, Network Infrastructure, Coverage and
Connectivity}
\label{sec:sn-infrastructure}

Sensor nodes can communicate in different modes such as radio, diffuse
light, inductive or sound, which may influence on the type of protocols used to
communicate among the nodes, where they can be said to be in an organized
infrastructure or as an ad hoc way \cite{sn-intro01}.

When it comes to infrastructure, sensor networks differ in topology:
single-hop, star, networked stars, tree and graph are examples of these
topologies. Each of they topologies may influence on how data is collection
and routed to their data sink. While there are only two levels of node
connectivity in sensor networks deployed as single-hop and single start
topologies, networks whose topology is either tree, graph or interconnected
stars use communications among each of their nodes for data communication.

Last, but not least, the coverage of a sensor network is related to the
physical space that is ``visible" to the sensors, which is directly related to
the purpose of the network, its size in terms of the number of nodes that are
deployed and the sensor network lifetime. In this way, sensor devices can be
deployed in different areas and distributed in sparse or dense manners, where
the sensor network they support is providing the necessary data that
represents the coverage of the network. However, considering that some types of
sensor devices are very likely to regular hardware failures, projects have
provided redundant nodes \cite{sn-intro01} to provide better data coverage.

No matter the type of coverage provided by a sensor network, different levels
of coverage may also influence on the connectivity of the sensor nodes. In
order to provide connectivity to redundant nodes, for example, a physical 
connection must be available at all times. On the other hand, sensor nodes can
also communicate with other nodes in a sporadic way when they are designed to
be isolated with only communication activities to report the collected data to
their immediate peer (other nodes and/or data sink).

@@

The network size is related to the degree of coverage of a sensor network, and
it is measured by the number of devices deployed as nodes of the network.
\cite{sn-intro01} reports examples of sensor networks with sensor nodes ranging
from dozens, hundreds or even thousand of network nodes.

The application of the sensor network determines for how long the sensor network
may be exist. For example, the existence of an environmental sensor networks
\cite{sn-ex01} may be not determined since its the data collected from the
network are used for archival purposes, hence, without termination of the
execution of its sensing tasks. On the other hand, however, the lifetime of a
sensor network designed to count the number of people who entered in different
rooms of a building during an event has its life-time related to the period of
the event.

@@

\subsection{How the Collected Data is Used}
\label{sec:sn-data-purpose}

@@

Under these circumstances, the collected data needs first to face its journey
starting from its transmission from its producing device to the specified data
sink as specified by \cite{sn-storage01, sn-storage02, sn-storage03}.

@@

It is clear that the data flows from the sensor node to another depending on
the definition of the network. On the whole, the sensor network infrastructure
dictates where the data will flow to, as described in section
\ref{sec:sn-infrastructure}. \cite{sn-storage02} sees a wireless sensor
networks as a many-to-one data collection pattern, when its organization is
based on single-hop or star. On the other hand, when the data organization is
structured as a graph, tree, etc, the data flows from one intermediate sensor
node to another until it reaches its final storage node as described by
\cite{sn-storage01, sn-storage03}.

@@

In order to mitigate the problems related flooded nodes in in-network query
processing, the use of a Data-Centric storage strategy has been shown to be a
proven alternative for the purpose of energy consumption
\cite{sn-storage03, sn-storage01}.Its primary goal is to organize the
collected data based on a particular property, say location or event, in order
to be sent to individual storage nodes that relate to the observation. As the
main outcome of this strategy is to create network partitioning strategy, the
query processor would hit specific sensor nodes with the needed data and,
therefore, would decrease the load and maximize the capacity on the actual
sensor devices.

@@

\begin{itemize}
  \item \textbf{Tabular Data Model}: comma-separated values are immensely
  popular due to its simplicity. Although data is not index , it is an easy
  approach for data exchange among researchers \cite{sn-provenance};
  \item \textbf{Relational Data Model}: it is used in the majority of the
  projects. Most of them due to the easy access to database systems, with
  refactored versions that support the implementation of sensor networks and
  its requirements \cite{sn-db-tinydb};
  \item \textbf{Structured Data Model}: similar to the relational model,
  structured data models are becoming popular due to the Internet. The use of
  XML data \cite{xml} is not yet popular, but there are a few projects using
  it \cite{sn-xml-usage01}\cite{sn-xml-usage02}, implementing its query by
  using a middleware \cite{sn-xml-middleware, sn-xml-query-engines}.
\end{itemize}

@@

\subsection{Relational Data Model and its Query Mechanism}

@@

\subsection{Structured Data Model and its Query Mechanism}

@@

\section{Data Provenance: How Collected Data is Described?}
\label{sec:sn-provenance}
